---
title: "R Benchmarkings"
author: "Francisco Charte Ojeda"
date: "Friday, February 27, 2015"
output:
  html_document:
    keep_md: yes
---

This document compares the performance in doing a task by means of different approaches in R. For doing so, the `microbenchmark` package is used, measuring the time spent by each approach. The results are shown numerically and plotting them using `ggplot2`. The numeric table shows relative performances, with the best method as `1.0` and the others showing the number of times which they are worse than the former.

The goal is to elucidate which is the best method to accomplish a certain task.

Generating a data.frame containing character data with and without stringsAsFactors
=================

With this code I want to test the difference between using `stringsAsFactors = TRUE` versus `stringsAsFactors = FALSE` while creating a new data.frame.

```{r init, include=FALSE, echo = FALSE}
library(knitr)
library(microbenchmark)
library(ggplot2)
library(compiler)
library(Rcpp)
library(inline)

opts_chunk$set(concordance=TRUE, cache = TRUE, cache.path = 'DocumentName_cache/', fig.path='figure/')
```

```{r testStringAsFactors}
numElements <- 1e6
someStrings <- sapply(1:25, function(x) paste(sample(c(letters, LETTERS), 10, replace = TRUE), collapse = ""))

aNumericVector <- runif(numElements)
aStringVector <- sample(someStrings, numElements, replace = TRUE)
bStringVector <- sample(someStrings, numElements, replace = TRUE)

result <- microbenchmark(
    data.frame(aNumericVector, aStringVector, bStringVector, stringsAsFactors = TRUE),
    data.frame(aNumericVector, aStringVector, bStringVector, stringsAsFactors = FALSE)
)
```

```{r echo = FALSE}
levels(result$expr) <- c('stringsAsFactors=T', 'stringsAsFactors=F')
print(result, unit="relative")
autoplot(result)
```

Conclusion
----------------
Generating a `data.frame` containing character columns is quicker when `stringsAsFactors = FALSE` is used. Nonetheless, it may be taken into account that this option implies the use of more memory, as character strings are stored individually instead of as numeric values referencing the factor levels. For this same reason, further operations such as sorting by a character column can take more time (compared with sorting by a factor column).

Growing list vs preallocated list vs lapply
===============
With the code shown below I want to test the differences between creating a list growing it, preallocating the elements, and using the `lapply` function. 

```{r testListGeneration}
numElements <- 1e4

result <- microbenchmark(
  { v1 <- list() ; for(i in 1:numElements) v1[[i]] <- someStrings },
  { v2 <- vector('list', numElements) ; for(i in 1:numElements) v2[[i]] <- someStrings },
  { v3 <- lapply(1:numElements, function(i) someStrings)}
)
```

```{r echo = FALSE, dependson="testListGeneration"}
levels(result$expr) <- c('Empty list', 'Preallocated list', 'lapply')
print(result, unit="relative")
autoplot(result)
```

Conclusion
----------------
There is no doubt that growing the list as items are added is a bad idea, since this method is much slower than the other two. The differences between preallocating the list and then populating it with a `for` loop or generating it with the `lapply` function are not as large, but certainly `lapply` has the advantage.

The result should be the same while working with a vector or a data.frame, instead of a list.

$ vs [[ operator
==============
The `$` operator is constantly used in R code to access lists and data.frames elements by name. The operator `[` could be used to do the same task, using numeric indexes instead. Is there any performance difference between them?


```{r testListAccess}
aList <- list( a = 5, b = 'list', c = list(c1 = 25))

result <- microbenchmark(
  { c(aList$a, aList$b, aList$c$c1) }, 
  { c(aList[[1]], aList[[2]], aList[[2]][[1]]) }
)
```

```{r echo = FALSE, dependson="testListAccess"}
levels(result$expr) <- c('$ operator', '[[ operator')
print(result, unit="relative")
autoplot(result)
```

Conclusion
------------
Although the difference between the two operators is very tight, it should be taken into account if we use these operators inside a loop or any other repetitve structure. Multiply the small difference by the number of times the operator is used during the program execution to assess if the effort worth it. 

Comparison of two vector values
======

Assume that you want to know which items in a vector `v` (values) have higher values than the corresponding items (by position) in another vector `t` (threshold). The goal is setting to 0 those values. This is a task that can be accomplished in several ways, for instance:

```{r testCompareVectors}
fgen <- function() runif(numElements, 1, 10)
v <- fgen() 
t <- fgen()

result <- microbenchmark(
  { for(i in 1:length(v)) if(v[i] > t[i]) v[i] <- 0 },
  { v <- mapply(function(a,b) if(a > b) 0 else a, v, t) },
  { v[which(v > t)] <- 0 },
  { v[v > t] <- 0 },
  { v <- ifelse(v > t, 0, v) }
)
```

```{r echo = FALSE, dependson="testCompareVectors"}
levels(result$expr) <- c('for', 'mapply', 'which', 'v > t', 'ifelse')
print(result, unit="relative")
autoplot(result)
```

As can be seen, `mapply` produces the worst performance, followed by the `for` loop. The quickest way to do the work is almost the simplest one, using the `which` function. This function returns the indexes of elements affected, while  with the expression `v[v > t] <- 0` an array of the same length than `v` and `t` is obtained and all their elements are tested to see if they are `TRUE` or `FALSE` before the assignment.

Simple functions can be vectorized by means of the `Vectorize` function in the base R package. Let us see how this approach performs against the best one of the previous tests:

```{r testVectorize}
v <- fgen() 
t <- fgen()
f <- function(a, b) if(a > b) 0 else a
vf <- Vectorize(f)

result <- microbenchmark(
  { v[which(v > t)] <- 0 },
  { v <- vf(v, t) }
)
```

```{r echo = FALSE, dependson="testVectorize"}
levels(result$expr) <- c('which', 'Vectorize')
print(result, unit="relative")
autoplot(result)
```

Conclusion
--------------
When it comes to apply some change to those items in a vector that satisfy a certain restriction, it seems that firstly obtaining the indexes, with the `which` function, and then making the change is the most efficient way of those compared here.

R source code vs R compiled code vs C++ code
======

Sometimes it is not easy to translate a loop into a vectorized expression or a call to `apply`. For instance, this happens when the operation to be made in a cycle depens on the result of a previous iteration. In these cases the loop R function containing the loop can be translated to bytecode, by means of the `cmpfun` function of the `compiler` package. Another alternative would be implementing that loop in C++ taking advantage of the `Rcpp` package. But, is it worth it?

Let us compare the performance of the same task implemented as a R function, as a compiled R function and as a C++ function:

```{r testRCompiled}
numElements <- 1e5
v <- fgen() 
t <- fgen()

f <- function(v, t) for(i in 1:length(v)) if(v[i] > t[i]) v[i] <- 0
fc <- cmpfun(f)
cppFunction('
    void fCpp(NumericVector v, NumericVector t) {
      for(int i = 0; i < v.size(); i++)
         v[i] = v[i] > t[i] ? 0 : v[i];
    }
')

result <- microbenchmark(f(v, t), fc(v, t), fCpp(v, t))
```

```{r echo = FALSE, dependson="testRCompiled"}
levels(result$expr) <- c('R source', 'R compiled', 'Rcpp')
print(result, unit="relative")
autoplot(result)
```

As can be seen the C++ function, embedded into R code with the `cppFunction`, is considerably quicker than the other two alternatives. Even compiling to bytecode, without the effort of installing the `Rcpp` package, can be worth it.

Would be the C++ implementation of this task quicker than the `which` function based solution proposed in an earlier section? Let us see:

```{r testRcppvsWhich, dependson="testRCompiled"}
v <- fgen() 
t <- fgen()

cppFunction('
    void fCpp(NumericVector v, NumericVector t) {
      for(int i = 0; i < v.size(); i++)
         v[i] = v[i] > t[i] ? 0 : v[i];
    }
')

result <- microbenchmark(v[which(v > t)] <- 0, fCpp(v, t))
```

```{r echo = FALSE, dependson="testRcppvsWhich"}
levels(result$expr) <- c('which', 'Rcpp')
print(result, unit="relative")
autoplot(result)
```

Although the improvement provided by the C++ function over `which` is not impressive, certainly we can save some time if we are comfortable writing C++ code.
