---
title: "R Benchmarkings"
author: "Francisco Charte Ojeda"
date: "Friday, February 27, 2015"
output: html_document
---

This document compares the performance in doing a task by means of different approaches in R. For doing so, the `microbenchmark` package is used, measuring the time spent by each approach. The results are shown numerically and plotting them using `ggplot2`.

The goal is to elucidate which is the best method to accomplish a certain task.

Generating a data.frame containing character data with and without stringAsFactors
=================

With this code I want to test the difference between using `stringAsFactors = TRUE` versus `stringAsFactors = FALSE` while creating a new data.frame.

```{r init, include=FALSE, echo = FALSE}
library(knitr)
opts_chunk$set(
concordance=TRUE, cache = TRUE, cache.path = 'DocumentName_cache/', fig.path='figure/'
)
```

```{r testStringAsFactors}
library(microbenchmark)
library(ggplot2)

numElements <- 1E6
someStrings <- sapply(1:25, function(x) paste(sample(c(letters, LETTERS), 10, replace = TRUE), collapse = ""))

aNumericVector <- runif(numElements)
aStringVector <- sample(someStrings, numElements, replace = TRUE)
bStringVector <- sample(someStrings, numElements, replace = TRUE)

result <- microbenchmark(
    data.frame(aNumericVector, aStringVector, bStringVector, stringsAsFactors = TRUE),
    data.frame(aNumericVector, aStringVector, bStringVector, stringsAsFactors = FALSE)
)
levels(result$expr) <- c('stringsAsFactors=T', 'stringsAsFactors=F')
result
autoplot(result)
```

Conclusion
----------------
Generating a `data.frame` containing character columns is quicker when `stringsAsFactors = FALSE` is used. Nonetheless, it may be taken into account that this option implies the use of more memory, as character strings are stored individually instead of numeric values referencing the factor levels.
