---
title: "R Benchmarkings"
author: "Francisco Charte Ojeda"
date: "Friday, February 27, 2015"
output:
  html_document:
    keep_md: yes
---

This document compares the performance in doing a task by means of different approaches in R. For doing so, the `microbenchmark` package is used, measuring the time spent by each approach. The results are shown numerically and plotting them using `ggplot2`.

The goal is to elucidate which is the best method to accomplish a certain task.

Generating a data.frame containing character data with and without stringsAsFactors
=================

With this code I want to test the difference between using `stringsAsFactors = TRUE` versus `stringsAsFactors = FALSE` while creating a new data.frame.

```{r init, include=FALSE, echo = FALSE}
library(knitr)
opts_chunk$set(
concordance=TRUE, cache = TRUE, cache.path = 'DocumentName_cache/', fig.path='figure/'
)
```

```{r testStringAsFactors}
library(microbenchmark)
library(ggplot2)

numElements <- 1e6
someStrings <- sapply(1:25, function(x) paste(sample(c(letters, LETTERS), 10, replace = TRUE), collapse = ""))

aNumericVector <- runif(numElements)
aStringVector <- sample(someStrings, numElements, replace = TRUE)
bStringVector <- sample(someStrings, numElements, replace = TRUE)

result <- microbenchmark(
    data.frame(aNumericVector, aStringVector, bStringVector, stringsAsFactors = TRUE),
    data.frame(aNumericVector, aStringVector, bStringVector, stringsAsFactors = FALSE)
)
```

```{r echo = FALSE}
levels(result$expr) <- c('stringsAsFactors=T', 'stringsAsFactors=F')
result
autoplot(result)
```

Conclusion
----------------
Generating a `data.frame` containing character columns is quicker when `stringsAsFactors = FALSE` is used. Nonetheless, it may be taken into account that this option implies the use of more memory, as character strings are stored individually instead of as numeric values referencing the factor levels. For this same reason, further operations such as sorting by a character column can take more time (compared with sorting by a factor column).

Growing list vs preallocated list vs lapply
===============
With the code shown below I want to test the differences between creating a list growing it, preallocating the elements, and using the `lapply` function. 

```{r testListGeneration}
numElements <- 1e4

result <- microbenchmark(
  { v1 <- list() ; for(i in 1:numElements) v1[[i]] <- someStrings },
  { v2 <- vector('list', numElements) ; for(i in 1:numElements) v2[[i]] <- someStrings },
  { v3 <- lapply(1:numElements, function(i) someStrings)}
)
```

```{r echo = FALSE, dependson="testListGeneration"}
levels(result$expr) <- c('Empty list', 'Preallocated list', 'lapply')
result 
autoplot(result)
```

Conclusion
----------------
There is no doubt that growing the list as items are added is a bad idea, since this method is much slower than the other two. The differences between preallocating the list and then populating it with a `for` loop or generating it with the `lapply` function are not as large, but certainly `lapply` has the advantage.

The result should be the same while working with a vector or a data.frame, instead of a list.

$ vs [[ operator
==============
The `$` operator is constantly used in R code to access lists and data.frames elements by name. The operator `[` could be used to do the same task, using numeric indexes instead. Is there any performance difference between them?


```{r testListAccess}
aList <- list( a = 5, b = 'list', c = list(c1 = 25))

result <- microbenchmark(
  { c(aList$a, aList$b, aList$c$c1) }, 
  { c(aList[[1]], aList[[2]], aList[[2]][[1]]) }
)
```

```{r echo = FALSE, dependson="testListAccess"}
levels(result$expr) <- c('$ operator', '[[ operator')
result 
autoplot(result)
```

Conclusion
------------
Although the difference between the two operators is very tight, it should be taken into account if we use these operators inside a loop or any other repetitve structure. Multiply the small difference by the number of times the operator is used during the program execution to assess if the effort worth it. 


